---
title: "Diabetes Data Analysis"
author: "Zekarias Mengiste"
date: "2025-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(kableExtra)
```
## Introduction

This analysis explores a publicly available diabetes dataset from Kaggle, containing demographic and medical features such as glucose levels, blood pressure, BMI, and insulin levels. The primary objective is to summarize key numerical variables: Glucose level, BMI, Age, and Pregnancies.
Loading the Data

```{r}
# Load the dataset
diabetes <- read.csv("data/diabetes.csv")

# Display first few rows
head(diabetes)
```

## Summary of Descriptive Statistics

The table below presents descriptive statistics for Age, BMI, Glucose, and Pregnancies in the dataset.
```{r}
# Compute summary statistics
desc_stats <- diabetes %>%
  summarise(
    Mean_Glucose = mean(Glucose, na.rm = TRUE),
    Median_Glucose = median(Glucose, na.rm = TRUE),
    SD_Glucose = sd(Glucose, na.rm = TRUE),
    Min_Glucose = min(Glucose, na.rm = TRUE),
    Max_Glucose = max(Glucose, na.rm = TRUE),
    
    Mean_Age = mean(Age, na.rm = TRUE),
    Median_Age = median(Age, na.rm = TRUE),
    SD_Age = sd(Age, na.rm = TRUE),
    Min_Age = min(Age, na.rm = TRUE),
    Max_Age = max(Age, na.rm = TRUE),
    
    Mean_BMI = mean(BMI, na.rm = TRUE),
    Median_BMI = median(BMI, na.rm = TRUE),
    SD_BMI = sd(BMI, na.rm = TRUE),
    Min_BMI = min(BMI, na.rm = TRUE),
    Max_BMI = max(BMI, na.rm = TRUE),
    
    Mean_Pregnancies = mean(Pregnancies, na.rm = TRUE),
    Median_Pregnancies = median(Pregnancies, na.rm = TRUE),
    SD_Pregnancies = sd(Pregnancies, na.rm = TRUE),
    Min_Pregnancies = min(Pregnancies, na.rm = TRUE),
    Max_Pregnancies = max(Pregnancies, na.rm = TRUE)
  )

# Reshape data to display each variable in a separate row
desc_stats_long <- desc_stats %>%
  gather(key = "Statistic", value = "Value") %>%
  separate(Statistic, into = c("Variable", "Measure"), sep = "_") %>%
  spread(key = "Measure", value = "Value")

# Display the table with styling
desc_stats_long %>%
  kable("html", caption = "Descriptive Statistics of Key Variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  row_spec(0, background = "yellow")
```

### Table description (Key findings):

Age: Ranges from 21 to 81 years, with a mean of 33.24.
BMI: Varies from 0 to 67.1, with a mean of 31.99 (BMI of 0 may indicate missing data).
Glucose: Spans from 0 to 199, with a mean of 120.89, showing considerable variation.
Pregnancies: Ranges from 0 to 17, with a mean of 3.85.


# Distribution of Blood Glucose Levels
```{r}

ggplot(diabetes, aes(x=Glucose)) + 
  geom_histogram(binwidth=10, fill="steelblue", color="black", alpha=0.7) + 
  labs(title="Distribution of Blood Glucose Levels", x="Glucose Level", y="Frequency")

```

### Description of figure
The above histogram shows the distribution of blood glucose levels among participants in the study. The data appears to follow a roughly normal distribution, with most values concentrated between 80 and 150. There is a peak around 100-120, indicating that many individuals have glucose levels in this range.The distribution is slightly right-skewed, with some individuals having higher glucose levels up to 200.

